Cryptobenthic reef assemblages are more distinct across a 90 m depth gradient than 2,500 km of shallow marine habitat in the Hawaiian Archipelago

### 2024-10-18 manuscript.Rmd
tiny tweak
revision updates

---
title: 'Cryptobenthic reef assemblages are more distinct across a 90 m depth gradient than 2,500 km of shallow marine habitat in the Hawaiian Archipelago'
author:
  - Kaleonani K.C. Hurley:
      institute:
        - himb
      equal_contributor: yes
  - Mykle L. Hoban:
      institute:
        - himb
      equal_contributor: yes
      correspondence: yes
      email: "mhoban@hawaii.edu"
  - Kerry Reardon:
      institute: noaa
  - Derek J. Skillings:
      institute: 
        - uncg
  - Molly A. Timmers:
      institute:
        - himb
        - natgeo
  - Robert J. Toonen:
      institute: himb
institute:
  - himb: Hawai&#x02BB;i Institute of Marine Biology, University of Hawai&#x02BB;i, Mānoa
  - noaa: National Oceanic and Atmospheric Administration
  - uncg: 'University of North Carolina, Greensboro'
  - natgeo: National Geographic Pristine Seas
bibliography: references/citations.json
csl: references/style.csl
output:
  word_document:
    pandoc_args:
      - "--filter=resources/pandoc-crossref"
      - "--citeproc"
      - "--toc-depth=2"
      - "--lua-filter=resources/scholarly-metadata.lua"
      - "--lua-filter=resources/author-info-blocks.lua"
      - "--lua-filter=resources/abstract-section.lua"
      - "--reference-doc=resources/ref.docx"
      - "--lua-filter=resources/multirefs.lua"
    keep_md: true
tblPrefix: table
figPrefix:
  - "figure"
  - "figures"
chapters: True
chaptersDepth: 1
chapDelim: ""
header-includes:
  - \usepackage{bm}
  - \usepackage{upgreek}
multiref_no_duplicates: true
---
<!--word count: Proceedings B publishes articles up to 12 pages, including figures and tables, with no more than 10 000 words. The word count includes the title page, author names, abstract, figure legends, tables and the reference list. As a rough guide, 10 000 words = 12 pages without additional figures; 850 words = 1 page and 2-4 small figures = 1 page. For example, an acceptable paper could be made up of 9 000 words and 4 small figures or 8500 words, 2 large figures and 4 small figures.-->
### Keywords {-}
<div custom-style="mono">Mesophotic coral ecosystems, community structure, depth gradient, environmental drivers, Hawaii, Autonomous Reef Monitoring Structures</div>

@@

@@

We assessed variation among groups using `betadisper` and `adonis2`. These functions implement analysis of multivariate homogeneity of group dispersions (PERMDISP) and multivariate permutational analysis of variance (PERMANOVA) respectively. PERMDISP is used to assess variability within groups and PERMANOVA is used to assess differences among groups. Significance was assessed with permutation tests in both cases. Variation among island groups and sampling depths was further examined and visualized using canonical analysis of principal coordinates (CAP) based on discriminant analysis [@andersonCanonicalAnalysisPrincipal2003] as implemented in the `CAPdiscrim` function in `BiodiversityR` [@kindtTreeDiversityAnalysis2005]. To explore differences in community variability among groups (shallow/deep, O&#x02BB;ahu/all shallow samples), we compared Euclidean distances among samples in the CAP ordination using Welch two-sample t-tests, assuming unequal variances.

@@

<!-- TODO: update number of different types to reflect family/genus/species level rather than species/morphospecies level -->
The `r nsamples(cc$crabs_untransformed)` ARMS units deployed across the Hawaiian Islands (in `r format_range(cc$crab_data$depth)` m depth) yielded a total of `r num(sum(sample_sums(cc$crabs_untransformed)),1)` adult crabs. Units contained `r num(all_summary$abundance_mean,0.1)` $\pm$ `r num(all_summary$abundance_sd,0.1)` (mean $\pm$ SD) individuals, and abundance ranged from `r format_range(all_summary$abundance_range)`. Crabs were identified to `r sum(unlist(all_summary$types_breakdown))` distinct taxa (`r all_summary$types_breakdown$species` species-level identifications, `r all_summary$types_breakdown$genus` genus, `r all_summary$types_breakdown$family` family) encompassing `r all_summary$taxonomy_breakdown$genera` genera in `r all_summary$taxonomy_breakdown$families` families, which accounted for ~`r pc(all_summary$types_breakdown$species / hawaiian_crabs,0.1)` of the known crab diversity reported from Hawaiian waters for species-level IDs [@castroCatalogAnomuranBrachyuran2011]. The four most abundant species were: `r with(all_summary,top_list(top,top_n,top_units,fig="fig_crabs"))`. There were `r all_summary$singleton_count` singleton taxa (taxa occurring in a single unit across all sites), accounting for ~`r pc(all_summary$singleton_count/ntaxa(cc$crabs_untransformed),0.1)` of total recovered crab taxa. For a breakdown of specific crab taxa recovered by island and deployment depth, see [@tbl:species_table].

@@

@@

For the complete dataset, there was no significant difference in observed species richness (`r k_test(all_alpha$k_r_island)`; [@fig:fig_alpha[]{}a]) or effective Simpson diversity (`r k_test(all_alpha$k_s_island,plain=TRUE)`; [@fig:fig_alpha[]{}c]) between the MHI and NWHI. Richness was not significantly different between shallow and deep samples (`r k_test(all_alpha$k_r_depth,plain=TRUE)`; [@fig:fig_alpha[]{}b]) but effective Simpson diversity was (`r k_test(all_alpha$k_s_depth,plain=TRUE)`; [@fig:fig_alpha[]{}d]). When comparing diversity across sampling depths for only sites in the MHI, richness was not significantly different (`r k_test(all_alpha$k_r_depth_main,plain=TRUE)`; [@fig:fig_alpha_main[]{}a]) but effective Simpson diversity was (`r k_test(all_alpha$k_s_depth_main,plain=TRUE)`; [@fig:fig_alpha_main[]{}b]). Neither richness nor Simpson diversity were significantly different between shallow and deep when comparisons were constrained to the island of O&#x02BB;ahu (Richness: `r k_test(all_alpha$k_r_depth_oahu,plain=TRUE)`; Simpson: `r k_test(all_alpha$k_s_depth_oahu,plain=TRUE)`). When only shallow arms \< 30 m were considered, there were no significant differences in alpha diversity between the MHI and NWHI for either index (richness: `r k_test(shallow_alpha$k_r_island,plain=TRUE)`, [@fig:fig_alpha2[]{}a]; effective Simpson: `r k_test(shallow_alpha$k_s_island,plain=TRUE)`, [@fig:fig_alpha2[]{}b]).

@@

all_terms <- english::english(nrow(model_terms))
```
### Environmental drivers {-}
Environmental variables were strongly clustered according to island group (MHI vs NWHI) ([@fig:fig_pca]). The best dbRDA model (`r permanova(db_modl,stat="F")`) explained `r pc(RsquareAdj(db_modl)$adj.r.squared,0.01)` (adjusted $\mathrm{R}^2$) of the variability in crab communities across the archipelago ([@fig:fig_dbrda]). `r str_to_sentence(english(nterm))` of the `r str_to_lower(english(nrow(model_terms)))` environmental variables tested were significant and sufficiently uncorrelated. The two drivers contributing most to explained variation were `r drivers$term[1]` (F = `r num(drivers$stat[1],0.01)`) and `r drivers$term[2]` (F = `r num(drivers$stat[2],0.01)`) ([@tbl:rda_table]). These factors also produced the longest effect vectors in the dbRDA plot ([@fig:fig_dbrda]).

@@

anova_depth <- adonis2(cc$crab_dist ~ depth_zone,data=cc$crab_data,permutations=PERM)
bd_depth <- betadisper(cc$crab_dist,cc$crab_data$depth_zone)

@@

distance_t <- t.test(deep_dist,shallow_dist,alternative = "g",var.equal = FALSE)
oahu_t <- t.test(oahu_dist,shallow_dist,alternative = "g",var.equal = FALSE)
```
Shallow crab community structure was significantly different between island groups (MHI vs. NWHI) (`r permanova(anova_region,0.01)`) as well as among individual islands (`r permanova(anova_island,0.01)`; see [@fig:fig_pairwise] for results of pairwise comparisons). There was no significant difference in community dispersion between the MHI and NWHI (`r permdisp(bd_region)`). When mesophotic ($\geq$ 30 m) ARMS units were included, there were significant differences among depth zones (\< 30 m, 30 m, 60 m, 90 m) (`r permanova(anova_depth,0.01)`) and community dispersion (`r permdisp(bd_depth)`). However, community dispersion was not different among depth zones $\geq$ 30 m (`r permdisp(bd_deep)`). Although crab communities were significantly different among both depth zones and island groups for the complete dataset (`r permanova(anova_both,0.01)`), depth explained `r num(effect_ratio,0.1)` times more variability than island group (based on model $\mathrm{R}^2$).

@@

```{r fig-cap, fig.cap="CAP ordination of crab communities by island group and depth zone.", results="asis", fig.width=9, fig.height=6, dependson="cap-analysis"}
centroids <- pcap %>% group_by(sample_group) %>%
  summarise(LD1=mean(LD1),LD2=mean(LD2))
pcap <- pcap %>%
  mutate(
    oahu = factor(
      if_else(island == "Oahu","oahu","noahu"),
      levels = c("oahu","noahu")
    )
  )

ggplot(pcap,aes(x=LD1,y=LD2,fill=sample_group)) + 
  stat_ellipse(geom="polygon",level=0.95,alpha=0.1,show.legend = FALSE) +
  # geom_point(aes(shape=oahu),color="black",shape=21,size=5)  +
  geom_point(mapping=aes(shape=oahu),color="black",size=5)  +
  geom_label_repel(data=centroids,aes(label=sample_group), color="white", show.legend = FALSE) + 
  scale_shape_manual(values=c("oahu" = 23, "noahu" = 21),name="Island", labels=c("noahu"="Other islands","oahu"="O‘ahu")) +
  scale_fill_manual(values=sample_depth_pal,name="Sample group") +
  guides( 
    fill=guide_legend(override.aes=list(shape=21)), 
    shape=guide_legend(override.aes=list(fill="black"))
  ) +
  xlab(str_glue("Linear discriminant 1 [{pc(cap_var$x,0.1)}]")) +
  ylab(str_glue("Linear discriminant 2 [{pc(cap_var$y,0.1)}]")) +
  plotz_theme()

# ggplot(pcap,aes(x=LD1,y=LD2,fill=sample_group)) + 
#   stat_ellipse(geom="polygon",level=0.95,alpha=0.1,show.legend = FALSE) +
#   geom_point(color="black",shape=21,size=5)  +
#   geom_label_repel(data=centroids,aes(label=sample_group), color="white", show.legend = FALSE) + 
#   scale_fill_manual(values=sample_depth_pal,name="Sample group") +
#   xlab(str_glue("Linear discriminant 1 [{pc(cap_var$x,0.1)}]")) +
#   ylab(str_glue("Linear discriminant 2 [{pc(cap_var$y,0.1)}]")) +
#   plotz_theme()
cat("{#fig:fig_cap}")
```

@@

# Discussion {-}
The Hawaiian Archipelago is among the most remote oceanic island chains on the planet and encompasses a range of gradients (anthropogenic, biotic, abiotic), making it an ideal natural laboratory to study the impacts of various drivers of species assemblages across space. Human activities are considerably higher in the populated MHI than in the protected and remote NWHI [@selkoeEvaluatingAnthropogenicThreats2008; @selkoeMapHumanImpacts2009; @selkoeDNACoralReef2016]. Although we did not consider the effects of individual impacts here, an index representing *cumulative* anthropogenic impacts was among the top drivers of shallow brachyuran community structure, similar to findings in previous work examining genetic diversity as it relates to species richness and habitat area, quality, and stability across the archipelago [@selkoeDNACoralReef2016]. The next-most significant driver, chlorophyll-A, can be used as a proxy for primary productivity. Productivity gradients across the Hawaiian Archipelago coincide with differences in coral reef community composition [@selkoeDNACoralReef2016], endemic reef fish dominance [@friedlanderDominanceEndemicsReef2020], survival rates of top predators [@bakerEffectVariableOceanic2007], and island geomorphology [@goveNearislandBiologicalHotspots2016]. Thus, it is not surprising that we find the same top drivers of community structure for one of the primary taxonomic groups inhabiting coral reefs, the brachyuran crabs, across the Hawaiian archipelago. What is surprising---given the maximum vertical separation among shallow ARMS units of `r diff(range(cc$crab_data_shallow_unscaled$depth))` m---is the consistent significance of depth as a driver of community composition across these gradients.

@@

While depth was a significant driver of community composition among shallow-water ARMS units ([@fig:fig_dbrda]), its influence was even more striking when comparing shallow to mesophotic samples. Due to logistical constraints (and the limitations of satellite-derived data products), our environmental data only covers shallow-water deployments, but CAP ordinations show considerable differences between the grouping of all shallow communities and the units deployed $\geq$ 30 m at the single island of O&#x02BB;ahu ([@fig:fig_cap]). According to these analyses, there was more community variation among samples spanning 90 m of depth on a single island than across 2,500 km of shallow habitat experiencing multiple gradients of environmental, ecological, and anthropogenic influence. The ordination revealed two distinct clusters for mesophotic samples: one comprising the 30 m samples and the other the 60 m and 90 m samples. These two clusters coincide with what is regarded as the upper and lower mesophotic zones, respectively [@kahngCommunityEcologyMesophotic2010; @rochaMesophoticCoralEcosystems2018; but see @pyleMesophoticCoralEcosystems2019]. The upper mesophotic (~30--60 m) has often been considered an extension of shallow reefs while the lower mesophotic (~60--150 m) is primarily thought to constitute a unique community. This pattern has held true across various taxonomic groups [@lesserGlobalCommunityBreaks2019; @rochaMesophoticCoralEcosystems2018], but our data show more distinct separations among shallow, upper, and lower mesophotic depths (although 30 m appears to encompass a transition zone).

@@

# Acknowledgements
We thank the NOAA Coral Reef Conservation Program for supporting the deployment and recovery of the shallow ARMS. We thank the UH Mānoa Dive Safety Office (D. Pence, K. Stender, J. Jones, and C.J. Bradley) for installation and retrieval of the mesophotic ARMS. We specially thank K. Flanagan from the UH dive program, whose support was instrumental in the development of the mesophotic ARMS project, and J. Copus, who provided expert mesophotic field support; both were taken too soon. May their memories be a blessing. We thank the members of the ToBo lab, particularly M. Iacchei, C. Ka&#x02BB;apu-Lyons, and Z. Hee, for discussion, support, and their efforts in the disassembly and sorting of ARMS samples on retrieval. Finally, we are grateful to the two anonymous reviewers for constructive comments and suggestions on an earlier version of this paper. This is contribution #xxx from the Hawai&#x02BB;i Institute of Marine Biology and #xxx from the School of Ocean and Earth Science and Technology at the University of Hawai&#x02BB;i.

@@

### Processing of satellite-derived environmental data
Level-3 mapped (4 km grid-size) sea-surface temperature (SST) and chlorophyll-A (chl-A) data were retrieved from NASA's Physical Oceanography Distributed Active Archive Center [PO.DAAC; @nasaobpgMODISAquaGlobal2020] and Ocean Biology Distributed Active Archive Center [OB.DAAC; @nasagoddardspaceflightcenterModerateresolutionImagingSpectroradiometer2022] respectively. Monthly summary datasets were downloaded for the period of 2009--2016 and averaged over each of the ARMS unit soaking periods (e.g. 2009--2011, 2010--2012, etc.). Values (SST or chl-A) were extracted for the geographic coordinates of each sampling site from the averaged dataset for the relevant soaking period. Original raw monthly-average dataset names can be found in @tbl:nasa_table.

>>>>

---
title: 'Cryptobenthic reef assemblages are more distinct across a 90 m depth gradient than 2,500 km of shallow marine habitat in the Hawaiian Archipelago'
author:
  - Mykle L. Hoban:
      institute:
        - himb
      equal_contributor: yes
      correspondence: yes
      email: "mhoban@hawaii.edu"
  - Kaleonani K.C. Hurley:
      institute:
        - himb
      equal_contributor: yes
  - Kerry Reardon:
      institute: noaa
  - Derek J. Skillings:
      institute: 
        - uncg
  - Molly A. Timmers:
      institute:
        - himb
        - natgeo
  - Robert J. Toonen:
      institute: himb
institute:
  - himb: Hawai&#x02BB;i Institute of Marine Biology, University of Hawai&#x02BB;i, Mānoa
  - noaa: National Oceanic and Atmospheric Administration
  - uncg: 'University of North Carolina, Greensboro'
  - natgeo: National Geographic Pristine Seas
bibliography: references/citations.json
csl: references/style.csl
output:
  word_document:
    pandoc_args:
      - "--filter=resources/pandoc-crossref"
      - "--citeproc"
      - "--toc-depth=2"
      - "--lua-filter=resources/scholarly-metadata.lua"
      - "--lua-filter=resources/author-info-blocks.lua"
      - "--lua-filter=resources/abstract-section.lua"
      - "--reference-doc=resources/ref.docx"
      - "--lua-filter=resources/multirefs.lua"
    keep_md: true
tblPrefix: table
figPrefix:
  - "figure"
  - "figures"
chapters: True
chaptersDepth: 1
chapDelim: ""
header-includes:
  - \usepackage{bm}
  - \usepackage{upgreek}
multiref_no_duplicates: true
---
<!--word count: Proceedings B publishes articles up to 12 pages, including figures and tables, with no more than 10 000 words. The word count includes the title page, author names, abstract, figure legends, tables and the reference list. As a rough guide, 10 000 words = 12 pages without additional figures; 850 words = 1 page and 2-4 small figures = 1 page. For example, an acceptable paper could be made up of 9 000 words and 4 small figures or 8500 words, 2 large figures and 4 small figures.-->
### Keywords {-}
<div custom-style="mono">Mesophotic coral ecosystems, community structure, depth gradient, environmental drivers, Hawaii, Autonomous Reef Monitoring Structures</div>

@@

# do an english comma-separated list
format_list <- function(l) {
  l[length(l)] <- str_c("and ",l[length(l)])
  if (length(l) > 2) { 
    str_c(l,collapse=", ")
  } else {
    str_c(l,collapse=" ")
  }
}

@@

We assessed variation among groups using `betadisper` and `adonis2`. These functions implement analysis of multivariate homogeneity of group dispersions (PERMDISP) and multivariate permutational analysis of variance (PERMANOVA) respectively. PERMDISP is used to assess variability within groups and PERMANOVA is used to assess differences among groups. Significance was assessed with permutation tests in both cases. We used PERMANOVA to make the following comparisons: island group (MHI vs NWHI, shallow samples only), island (shallow samples only), depth zone only (all samples), depth zone only (O&#x02BB;ahu samples), island group and depth zone (all samples). Variation among island groups and sampling depths was further examined and visualized using canonical analysis of principal coordinates (CAP) based on discriminant analysis [@andersonCanonicalAnalysisPrincipal2003] as implemented in the `CAPdiscrim` function from `BiodiversityR` [@kindtTreeDiversityAnalysis2005]. To explore differences in community variability among groups (shallow/deep, O&#x02BB;ahu/all shallow samples), we compared Euclidean distances among samples in the CAP ordination using Welch two-sample t-tests, assuming unequal variances.

@@

<!-- TODO: update number of different types to reflect family/genus/species level rather than species/morphospecies level -->
The `r nsamples(cc$crabs_untransformed)` ARMS units deployed across the Hawaiian Islands (in `r format_range(cc$crab_data$depth)` m depth) yielded a total of `r num(sum(sample_sums(cc$crabs_untransformed)),1)` adult crabs. Units contained `r num(all_summary$abundance_mean,0.1)` $\pm$ `r num(all_summary$abundance_sd,0.1)` (mean $\pm$ SD) individuals, and abundance ranged from `r format_range(all_summary$abundance_range)`. Crabs were identified to `r sum(unlist(all_summary$types_breakdown))` distinct taxa (`r all_summary$types_breakdown$species` species-level identifications, `r all_summary$types_breakdown$genus` genus, `r all_summary$types_breakdown$family` family) encompassing `r all_summary$taxonomy_breakdown$genera` genera in `r all_summary$taxonomy_breakdown$families` families, which accounted for ~`r pc(all_summary$types_breakdown$species / hawaiian_crabs,0.1)` of the known crab diversity reported from Hawaiian waters for species-level IDs [@castroCatalogAnomuranBrachyuran2011]. The four most abundant species were: `r with(all_summary,top_list(top,top_n,top_units,fig="fig_crabs"))`. There were `r all_summary$singleton_count` singleton taxa (taxa occurring in a single unit across all sites), accounting for ~`r pc(all_summary$singleton_count/ntaxa(cc$crabs_untransformed),0.1)` of total recovered crab taxa. For a breakdown of specific crab taxa recovered by island and deployment depth, see [@tbl:species_table_s].

@@

### Comparisons {-}
```{r depth-ranges, results="hide", dependson="taxonomy-validation"}
f <- cc$crabs %>%
  psmelt() %>%
  group_by(depth_zone) %>%
  filter(Abundance > 0) %>%
  ungroup() %>%
  group_by(family) %>%
  filter(n_distinct(shallow_deep) == 1) %>%
  ungroup()

fam <- f %>%
  distinct(family,shallow_deep) %>%
  count(shallow_deep) %>%
  deframe()

deep_fams <- f %>%
  filter(shallow_deep == "deep") %>%
  pull(family) %>%
  unique() %>%
  sort()

g <- cc$crabs %>%
  psmelt() %>%
  filter(!is.na(genus)) %>%
  group_by(depth_zone) %>%
  filter(Abundance > 0) %>%
  ungroup() %>%
  group_by(genus) %>%
  filter(n_distinct(shallow_deep) == 1) %>%
  ungroup()

gen <- g %>%
  distinct(genus,shallow_deep) %>%
  count(shallow_deep) %>%
  deframe()

deep_gens <- g %>%
  filter(shallow_deep == "deep") %>%
  pull(genus) %>%
  unique() %>%
  sort()

s <- cc$crabs %>%
  psmelt() %>%
  group_by(depth_zone) %>%
  filter(Abundance > 0) %>%
  ungroup() %>%
  group_by(species) %>%
  filter(n_distinct(shallow_deep) == 1) %>%
  ungroup()

spp <- s %>%
  distinct(species,shallow_deep) %>%
  count(shallow_deep) %>%
  deframe()

deep_spp <- s %>%
  filter(shallow_deep == "deep") %>%
  pull(species) %>%
  unique() %>%
  sort()
```
Out of the `r all_summary$taxonomy_breakdown$families` families recovered, `r fam['shallow']` were exclusive to shallow ARMS recovered from `r dr(cc$crabs_shallow_untransformed)` m and `r fam['deep']` were exclusive to units $\geq$ 30 m (`r format_list(deep_fams)`; [@tbl:family_depth_table_s]). `r str_to_sentence(english(gen['shallow']))` genera were found exclusively shallow and `r gen['deep']` exclusively deep (`r format_list(deep_gens)`; [@tbl:genus_depth_table_s]). `r str_to_sentence(english(spp['shallow']))` distinct taxa (including those collapsed to genus or family) were exclusive to shallow ARMS units and `r spp['deep']` were found only deep (`r format_list(deep_spp)`; [@tbl:species_depth_table_s]).

@@

For the complete dataset, there was no significant difference in observed species richness (`r k_test(all_alpha$k_r_island)`; [@fig:fig_alpha[]{}a]) or effective Simpson diversity (`r k_test(all_alpha$k_s_island,plain=TRUE)`; [@fig:fig_alpha[]{}c]) between the MHI and NWHI. Richness was not significantly different between shallow and deep samples (`r k_test(all_alpha$k_r_depth,plain=TRUE)`; [@fig:fig_alpha[]{}b]) but effective Simpson diversity was (`r k_test(all_alpha$k_s_depth,plain=TRUE)`; [@fig:fig_alpha[]{}d]). When comparing diversity across sampling depths for only sites in the MHI, richness was not significantly different (`r k_test(all_alpha$k_r_depth_main,plain=TRUE)`; [@fig:fig_alpha_main[]{}a]) but effective Simpson diversity was (`r k_test(all_alpha$k_s_depth_main,plain=TRUE)`; [@fig:fig_alpha_main[]{}c]). Neither richness nor Simpson diversity were significantly different between shallow and deep when comparisons were constrained to the island of O&#x02BB;ahu (Richness: `r k_test(all_alpha$k_r_depth_oahu,plain=TRUE)`, [@fig:fig_alpha_main[]{}b]; Simpson: `r k_test(all_alpha$k_s_depth_oahu,plain=TRUE)`, [@fig:fig_alpha_main[]{}d]). When only shallow arms \< 30 m were considered, there were no significant differences in alpha diversity between the MHI and NWHI for either index (richness: `r k_test(shallow_alpha$k_r_island,plain=TRUE)`, [@fig:fig_alpha2[]{}a]; effective Simpson: `r k_test(shallow_alpha$k_s_island,plain=TRUE)`, [@fig:fig_alpha2[]{}b]).

@@

all_terms <- english::english(nrow(model_terms))
```
### Environmental drivers {-}
Environmental variables were strongly clustered according to island group (MHI vs NWHI) ([@fig:fig_pca]). The best dbRDA model (`r permanova(db_modl,stat="F")`) explained `r pc(RsquareAdj(db_modl)$adj.r.squared,0.01)` (adjusted $\mathrm{R}^2$) of the variability in crab communities across the archipelago ([@fig:fig_dbrda]). `r str_to_sentence(english(nterm))` of the `r str_to_lower(english(nrow(model_terms)))` environmental variables tested were significant and sufficiently uncorrelated. The two drivers contributing most to explained variation were `r drivers$term[1]` (F = `r num(drivers$stat[1],0.01)`) and `r drivers$term[2]` (F = `r num(drivers$stat[2],0.01)`) ([@tbl:rda_table_s]). These factors also produced the longest effect vectors in the dbRDA plot ([@fig:fig_dbrda]).

@@

anova_depth <- adonis2(cc$crab_dist ~ depth_zone,data=cc$crab_data,permutations=PERM)
anova_depth_oahu <- adonis2(cc$crab_dist_oahu ~ depth_zone,data=sample_tibble(cc$crabs_oahu),permutations=PERM)
anova_depth2_oahu <- adonis2(cc$crab_dist_oahu ~ shallow_deep,data=sample_tibble(cc$crabs_oahu),permutations=PERM)
bd_depth <- betadisper(cc$crab_dist,cc$crab_data$depth_zone)

@@

distance_t <- t.test(deep_dist,shallow_dist,alternative = "g",var.equal = FALSE)
oahu_t <- t.test(oahu_dist,shallow_dist,alternative = "g",var.equal = FALSE)
```
Shallow crab community structure was significantly different between island groups (MHI vs. NWHI) (`r permanova(anova_region,0.01)`) as well as among individual islands (`r permanova(anova_island,0.01)`; see [@fig:fig_pairwise] for results of pairwise comparisons). There was no significant difference in community dispersion between the MHI and NWHI (`r permdisp(bd_region)`). When mesophotic ($\geq$ 30 m) ARMS units were included, there were significant differences among depth zones (\< 30 m, 30 m, 60 m, 90 m) both for all samples (`r permanova(anova_depth,0.01)`) and samples restricted to O&#x02BB;ahu (`r permanova(anova_depth_oahu,0.01)`), and community dispersion (`r permdisp(bd_depth)`). However, community dispersion was not different among depth zones $\geq$ 30 m (`r permdisp(bd_deep)`). Although crab communities were significantly different among both depth zones and island groups for the complete dataset (`r permanova(anova_both,0.01)`), depth explained `r num(effect_ratio,0.1)` times more variability than island group (based on model $\mathrm{R}^2$).

@@

```{r fig-cap, fig.cap="CAP ordination of crab communities by island group and depth zone.", results="asis", fig.width=9, fig.height=6, dependson="cap-analysis"}
centroids <- pcap %>% group_by(sample_group) %>%
  summarise(LD1=mean(LD1),LD2=mean(LD2))
pcap <- pcap %>%
  mutate(
    oahu = factor(
      # if_else(island == "Oahu","oahu","noahu"),
      case_when(
        island == "Oahu" ~ "oahu",
        island == "French Frigate Shoals" ~ "ffs",
        .default = "noahu"
      ),
      levels = c("oahu","ffs","noahu")
    )
  )

ggplot(pcap,aes(x=LD1,y=LD2,fill=sample_group)) + 
  stat_ellipse(geom="polygon",level=0.95,alpha=0.1,show.legend = FALSE) +
  # geom_point(aes(shape=oahu),color="black",shape=21,size=5)  +
  geom_point(mapping=aes(shape=oahu),color="black",size=5)  +
  geom_label_repel(data=centroids,aes(label=sample_group), color="white", show.legend = FALSE) + 
  scale_shape_manual(values=c("oahu" = 23, "ffs" = 22, "noahu" = 21),name="Island", labels=c("noahu"="Other islands","oahu"="O‘ahu","ffs"="Lalo (French Frigate Shoals)" )) +
  scale_fill_manual(values=sample_depth_pal,name="Sample group") +
  guides( 
    fill=guide_legend(override.aes=list(shape=21)), 
    shape=guide_legend(override.aes=list(fill="black"))
  ) +
  xlab(str_glue("Linear discriminant 1 [{pc(cap_var$x,0.1)}]")) +
  ylab(str_glue("Linear discriminant 2 [{pc(cap_var$y,0.1)}]")) +
  plotz_theme()

# centroids <- pcap %>% group_by(sample_group) %>%
#   summarise(LD1=mean(LD1),LD2=mean(LD2))
# pcap <- pcap %>%
#   mutate(
#     oahu = factor(
#       if_else(island == "Oahu","oahu","noahu"),
#       levels = c("oahu","noahu")
#     )
#   )

# ggplot(pcap,aes(x=LD1,y=LD2,fill=sample_group)) + 
#   stat_ellipse(geom="polygon",level=0.95,alpha=0.1,show.legend = FALSE) +
#   # geom_point(aes(shape=oahu),color="black",shape=21,size=5)  +
#   geom_point(mapping=aes(shape=oahu),color="black",size=5)  +
#   geom_label_repel(data=centroids,aes(label=sample_group), color="white", show.legend = FALSE) + 
#   scale_shape_manual(values=c("oahu" = 23, "noahu" = 21),name="Island", labels=c("noahu"="Other islands","oahu"="O‘ahu")) +
#   scale_fill_manual(values=sample_depth_pal,name="Sample group") +
#   guides( 
#     fill=guide_legend(override.aes=list(shape=21)), 
#     shape=guide_legend(override.aes=list(fill="black"))
#   ) +
#   xlab(str_glue("Linear discriminant 1 [{pc(cap_var$x,0.1)}]")) +
#   ylab(str_glue("Linear discriminant 2 [{pc(cap_var$y,0.1)}]")) +
#   plotz_theme()
cat("{#fig:fig_cap}")
```

@@

# Discussion {-}
  The Hawaiian Archipelago is among the most remote oceanic island chains on the planet and encompasses a range of gradients (anthropogenic, biotic, abiotic), making it an ideal natural laboratory to study the impacts of various drivers of species assemblages across space. Human activities are considerably higher in the populated MHI than in the protected and remote NWHI [@selkoeEvaluatingAnthropogenicThreats2008; @selkoeMapHumanImpacts2009; @selkoeDNACoralReef2016]. Although we did not consider the effects of individual impacts here, an index representing *cumulative* anthropogenic impacts was among the top drivers of shallow brachyuran community structure, similar to findings in previous work examining genetic diversity as it relates to species richness and habitat area, quality, and stability across the archipelago [@selkoeDNACoralReef2016]. The next-most significant driver, chlorophyll-A, can be used as a proxy for primary productivity. Productivity gradients across the Hawaiian Archipelago coincide with differences in coral reef community composition [@selkoeDNACoralReef2016], endemic reef fish dominance [@friedlanderDominanceEndemicsReef2020], survival rates of top predators [@bakerEffectVariableOceanic2007], and island geomorphology [@goveNearislandBiologicalHotspots2016]. Although not a significant factor in our model, coral cover and species composition are thought to affect patterns of benthic assemblages. Due to the limited availability of datasets covering our entire study area, our model only included coral cover at the island scale, but previous work has shown that coral species composition does not vary considerably across the archipelago [@griggCommunityStructureSuccession1983]. Crab communities from Lalo (French Frigate Shoals), which was found to harbor the highest coral species diversity [@griggCommunityStructureSuccession1983], did not group separately from the rest of our samples ([@fig:fig_cap]). Given the prevalence in our model of significant factors known to be important influencers of marine communities in Hawai&#x02BB;i, it is not surprising that we find the same top drivers of community structure for brachyuran crabs, one of the primary taxonomic groups inhabiting coral reefs. What is surprising---given the maximum vertical separation among shallow ARMS units of `r diff(range(cc$crab_data_shallow_unscaled$depth))` m---is the consistent significance of depth as a driver of community composition across these gradients.

@@

While depth was a significant driver of community composition among shallow-water ARMS units ([@fig:fig_dbrda; @tbl:rda_table_s]), its influence was even more striking when comparing shallow to mesophotic samples. Due to logistical constraints (and the limitations of satellite-derived data products), our environmental data only covers shallow-water deployments, but CAP ordinations show considerable differences between the grouping of all shallow communities and the units deployed $\geq$ 30 m at the single island of O&#x02BB;ahu ([@fig:fig_cap]). According to these analyses, there was more community variation among samples spanning 90 m of depth on a single island than across 2,500 km of shallow habitat experiencing multiple gradients of environmental, ecological, and anthropogenic influence. The ordination revealed two distinct clusters for mesophotic samples: one comprising the 30 m samples and the other the 60 m and 90 m samples. These two clusters coincide with what is regarded as the upper and lower mesophotic zones, respectively [@kahngCommunityEcologyMesophotic2010; @rochaMesophoticCoralEcosystems2018; but see @pyleMesophoticCoralEcosystems2019]. The upper mesophotic (~30--60 m) has often been considered an extension of shallow reefs while the lower mesophotic (~60--150 m) is primarily thought to constitute a unique community. This pattern has held true across various taxonomic groups [@lesserGlobalCommunityBreaks2019; @rochaMesophoticCoralEcosystems2018], but our data show more distinct separations among shallow, upper, and lower mesophotic depths (although 30 m appears to encompass a transition zone).

@@

# Acknowledgements
We thank the NOAA Coral Reef Conservation Program for supporting the deployment and recovery of the shallow ARMS. We thank the UH Mānoa Dive Safety Office (D. Pence, K. Stender, J. Jones, and C.J. Bradley) for installation and retrieval of the mesophotic ARMS. We specially thank K. Flanagan from the UH dive program, whose support was instrumental in the development of the mesophotic ARMS project, and J. Copus, who provided expert mesophotic field support; both were taken too soon. May their memories be a blessing. We thank the members of the ToBo lab, particularly M. Iacchei, C. Ka&#x02BB;apu-Lyons, and Z. Hee, for discussion, support, and their efforts in the disassembly and sorting of ARMS samples on retrieval. Finally, we are grateful to two anonymous reviewers for constructive comments and suggestions on an earlier version of this paper. This is contribution #xxx from the Hawai&#x02BB;i Institute of Marine Biology and #xxx from the School of Ocean and Earth Science and Technology at the University of Hawai&#x02BB;i.

@@

### Processing of satellite-derived environmental data
Level-3 mapped (4 km grid-size) sea-surface temperature (SST) and chlorophyll-A (chl-A) data were retrieved from NASA's Physical Oceanography Distributed Active Archive Center [PO.DAAC; @nasaobpgMODISAquaGlobal2020] and Ocean Biology Distributed Active Archive Center [OB.DAAC; @nasagoddardspaceflightcenterModerateresolutionImagingSpectroradiometer2022] respectively. Monthly summary datasets were downloaded for the period of 2009--2016 and averaged over each of the ARMS unit soaking periods (e.g. 2009--2011, 2010--2012, etc.). Values (SST or chl-A) were extracted for the geographic coordinates of each sampling site from the averaged dataset for the relevant soaking period. Original raw monthly-average dataset names can be found in @tbl:nasa_table_s.
